### 2018-07-30 chapters/03-translator/forward-trans.tex
Implemented all suggestions by Howard

The parsed tree generated by the \gls*{pom}-tagger is not a mathematical expression tree. The \gls*{pom} project aims to disambiguate mathematical \LaTeX{} expressions and generates an expression tree. However, in the current state, many expressions cannot be disambiguated yet. In consequence, the \gls*{pom}-tagger generates a raw parsed tree where each token in the \LaTeX{} expression is a node in the tree. We call this parsed tree the \gls*{pom-pt}.

{\sloppy The overall forward translation process is explained in figure~\ref{fig:forward-trans}. All translation patterns and related information are stored in the DLMF/DRMF tables. These tables are converted by the \verb|lexicon-creator| to the \verb|DLMF-macros-lexicon| lexicon file. Together with the \verb|global|-\verb|lexicon| file, the \gls*{pom-pt} will be created by the \gls*{pom}-tagger. The \verb|latex-converter| takes a string representation of a semantic \LaTeX{} expression and uses the \gls*{pom} engine as well as our \verb|Translator| to create an appropriate string representation for a specified \gls*{cas}.}

@@

\subsection{Analyzing the PoM-Parsed Tree}\label{subsec:analyze-mlp}
Since the \gls*{bnf} does not define rules for semantic macros, each argument of the semantic macro and each $@$ symbol are following siblings of the semantic macro node. That is the reason, why we stored the number of parameters, variables and $@$ symbols in the lexicon files. Otherwise, the translator could not find the end of a semantic macro in the \gls*{pom-pt}.

@@

Figure~\ref{fig:syntax-tree-usecase} visualizes the \gls*{pom-pt} of the Jacobi polynomial example from table~\ref{tab:JacobiP-usecase}. Because of the differences to expression trees, a backward conversion of the \gls*{pom-pt} to a string representation can be difficult, especially for finding necessary or unnecessary parentheses. Therefore we create the \gls*{teo}.

With these tools, we can translate a \LaTeX{} expression by translating the \gls*{pom-pt} node by node and perform group or reordering operations for some special cases. The algorithm is realized in a simple recursive structure. The algorithm~\ref{alg:simple-translation} explains the translation process. Whenever the algorithm finds a leaf, it can translate this single term. If the node is not a leaf, it starts to translate all children of the node recursively.

\begin{algorithm}[H]
\caption{Abstract translation algorithm to translate MLP-Parse trees.}\label{alg:translation}
	\begin{algorithmic}[1]
	\Require Root $r$ of a POM-Parse tree $T$. List \textit{following\_siblings} with the following siblings of $r$. The list can be empty.
	\Procedure{abstract\_translator}{$r$, \textit{following\_siblings}}
	\If{$r$ is leaf}
		\State {\scriptsize TRANSLATE\_LEAF}($r$, \textit{following\_siblings});\label{line:transLeaf}
	\Else
		\State \textit{siblings} $ = r$.getChildren(); \Comment{\textit{siblings} is a list of children}
		\State {\scriptsize ABSTRACT\_TRANSLATOR}(\textit{siblings}.removeFirst(), \textit{siblings});\label{line:transSeq}
	\EndIf
	\If{\textit{following\_siblings} is not empty}\label{line:ifNotEmpty}
		\State $r =$ \textit{following\_siblings}.removeFirst();
		\State {\scriptsize ABSTRACT\_TRANSLATOR}($r$, \textit{following\_siblings});
	\EndIf
	\EndProcedure
	\end{algorithmic}
\end{algorithm}
\vspace{-20px}

This approach is not completely feasible since the algorithm needs to look ahead and check the following siblings in some cases, e.g., in the case of a semantic macro with arguments (see figure~\ref{fig:syntax-tree-usecase}). The {\footnotesize \verb|TRANSLATE_LEAF|} function needs to know the following siblings of the current node. Algorithm~\ref{alg:translation} explains a more detailed but abstract version of the final algorithm of the presented translation tool.

If the root $r$ is a leaf, it still can be translated as a leaf. Eventually, some of the following siblings are needed to translate $r$. The list of \textit{following\_siblings} in line~\ref{line:transLeaf} might be reduced to avoid multiple translations for one node. If $r$ is not a leaf, it contains one or more children. Therefore, we can call the {\footnotesize \verb|ABSTRACT_TRANSLATOR|} recursively for the children. Once we have translated $r$, we can go a step further and translate the next node. Line~\ref{line:ifNotEmpty} checks if there are following siblings left and calls the {\footnotesize \verb|ABSTRACT_TRANSLATOR|} recursively in that case. Translated expressions are stored by the \gls*{teo} object. The detailed translation process is not included in the algorithm.

Algorithm~\ref{alg:translation} is a simplified version of the translator process. The main progress happens in \cref{line:transLeaf,line:transSeq}. There are several cases what $r$ can be. Table~\ref{tab:allTypesTable} on page \pageref{tab:allTypesTable} gives an overview of all different types in a \gls*{pom-pt}. A more detailed explanation of the types can be found in~\parencite{POM-Tagger}.

The \gls*{bnf} grammar defines some basic grammatical rules for generic \LaTeX{} macros, such as for \verb|\frac|, \verb|\sqrt|. Therefore, there is a hierarchical structure for those symbols, similar to the structure in expression trees. As already mentioned, some of these types can be translated directly, such as Greek letters, while others are more complex, such as semantic \LaTeX{} macros. Therefore, the translators delegate the translation to specialized subtranslators. This delegation process is implemented in \cref{line:transLeaf,line:transSeq} of algorithm~\ref{alg:translation}. The Subsection~\ref{sec:subtranslators} discusses these classes in more detail.

This approach seems to work for most of semantic \LaTeX{} expressions. However, there is one problem with this straight forward approach which was mentioned shortly mentioned in section~\ref{sec:problems}. There are many different types of notations used to represent formulae. Table~\ref{tab:notations} illustrates the simple expression $(a+b)x$ in different notations. The \gls*{pol}\footnote{Also known as \textit{Warsaw Notation} or \textit{prefix notation}} (hereafter called prefix notation) places the operator to the left of/before its operands. The \gls*{rpol}\footnote{Also known as \textit{postfix notation}} (hereafter called postfix notation) does the opposite and places the operator to the right of/after its operands. The infix notation is commonly used in arithmetic and places the operator between their operands, which only makes sense as long as the operator is a binary operator.

\begin{wrapfigure}{r}{0.4\textwidth}
\vspace{-22pt}
\begin{minipage}{0.4\textwidth}
\center
\begin{tabular}{lc}
	\hline
	Notation & Expression \\
	\hline
	Infix & $(a+b) \cdot x$\\
	Prefix & $\cdot + a\ b\ x$\\
	Postfix & $a\ b + x\ \cdot$\\
	Functional & $\cdot(+(a, b), x)$\\
	\hline
\end{tabular}
\vspace{-5pt}
\caption{The mathematical expression '$(a+b) \cdot x$' in infix, prefix, postfix and functional notation.}
\label{tab:notations}
\end{minipage}
\vspace{-15pt}
\end{wrapfigure}

@@

Most programming languages (and \gls*{cas} as well) internally use prefix or postfix notation and do not mix the notations in one expression, since it is more convenient to parse expressions in uniform notations. However, the common practice in science is to use mixed notations in expressions. Since the \gls*{pom} has rarely implemented mathematical grammatical rules yet, it takes the input as it is and does not build an expression tree. Therefore, it parses all four examples from table~\ref{tab:notations} to four different \gls*{pom-pt}s rather than to one unique expression tree. In general, this is not a problem for our translation process since most \gls*{cas} are familiar with the most common notations. Therefore, the translator does not need to know that $a$ and $b$ are the operands of the binary operator '$+$' in '$a+b$.' The translator could simply translate the symbols in '$a+b$' in the same order as they appear in the expression and the \gls*{cas} would understand it. However, this simple approach generates two problems.
\begin{enumerate}%[label=\arabic*)]
\item \label{prob:1} The translated expression is only syntactically correct if the input expression was syntactically correct.
\item \label{prob:2} We cannot translate expressions to a \gls*{cas} which uses a non-standard notation.
\end{enumerate}

@@

Nonetheless, changing a notation could also solve ambiguities in some situations. Consider the two ambiguous examples in table~\ref{tab:amb_ex}. While a scientist would probably just ask for the right interpretation of the first example, \Maple{} automatically computes the first interpretation. On the other hand, \LaTeX{} automatically disambiguate the first example by only recognizing the very next element (single symbols or sequence in curly brackets) for the superscript and therefore displays the second interpretation. The second example is already interpreted as the double factorial function of $n$, since this notation is the standard interpretation in science. We wrote the second interpretation as the standard way in science to make it even more obvious. However, surprisingly, \Maple{} computes the first interpretation again rather than the common standard interpretation.
\begin{table}[ht]
\centering
\begin{tabular}{lccc}
	\hline
	& Text Format Expression & First Interpretation & Second Interpretation\\
	\hline
	1: & \rule{0pt}{0.9\normalbaselineskip} $4\ \hat{\ }\ 2!$ & $4^{2!}$ & $4^2!$ \\
	2: & $n!!$ & $(n!)!$ & $(n)!!$\\
	\hline
\end{tabular}
\caption{Ambiguous examples of the factorial and double factorial function. One expression in a text format can be interpreted in different ways.}
\label{tab:amb_ex}
\end{table}

In most cases, parentheses can be used to disambiguate expressions. We used them in table~\ref{tab:amb_ex} to clarify the different interpretations in example 2. But sometimes, even parentheses cannot solve a mistaken computation. For example, there is no way to add parentheses to force \Maple{} to compute $n!!$ as the double factorial function. Even $(n)!!$ will be interpreted as $(n!)!$. Rather than using the exclamation mark in \Maple, one could also use the functional notation. For example, the interpretations $(2!)!$ and $(2)!!$ can be distinguished in \Maple{} by using \verb|factorial(factorial(2))| and \verb|doublefactorial(2)| respectively. We define the translations as follows:

@@

An exponent will be translated and shifted to the end of the translated semantic macro, since it is more common to write the exponent after the arguments of a function in \gls*{cas}. Therefore, the function translates and stores the exponent in line~\ref{line:next_caret}. One could ask what happens when there is an exponent given before and after the arguments. The \verb|MacroTranslator| only translates the following siblings until each argument is translated. The first exponent will be shifted to the end. If right after the translated macro (with all arguments) follows another exponent, we interpret it as another exponent for the whole previous expression. In that case, it would be the macro with the first translated exponent. Table~\ref{tab:multi-expo} shows an example for the trigonometric cosine function with multiple exponents. 
\begin{table}[ht]
\centering
\begin{tabular}{lc}
\hline
Displayed As & \rule{0pt}{0.9\normalbaselineskip} $\cos^2(x)^2$ \\
Semantic \LaTeX & \verb|\cos^2@{x}^2| \\
Translated \Maple{} Expression & \verb|((cos(x))^(2))^2|\\
\hline
\end{tabular}
\caption{A trigonometric cosine function example with exponents before and after the argument.}
\label{tab:multi-expo}
\end{table}

>>>>

The parsed tree generated by the \gls*{pom}-tagger is not a mathematical expression tree. The \gls*{pom} project aims to disambiguate mathematical \LaTeX{} expressions and generates an expression tree. However, in the current state, many expressions cannot be disambiguated yet. In consequence, the \gls*{pom}-tagger generates a raw parsed tree where each token in the \LaTeX{} expression is a node in the tree. We call this parsed tree, the \gls*{pom-pt}.

{\sloppy The overall forward translation process is explained in~\Cref{fig:forward-trans}. All translation patterns and related information are stored in the DLMF/DRMF tables. These tables are converted by the \verb|lexicon-creator| to the \verb|DLMF-macros-lexicon| lexicon file. Together with the \verb|global|-\verb|lexicon| file, the \gls*{pom-pt} will be created by the \gls*{pom}-tagger. The \verb|latex-converter| takes a string representation of a semantic \LaTeX{} expression and uses the \gls*{pom} engine as well as our \verb|Translator| to create an appropriate string representation for a specified \gls*{cas}.}

@@

\subsection{Analyzing the PoM-Parsed Tree}\label{subsec:analyze-mlp}
Since the \gls*{bnf} does not define rules for semantic macros, each argument of the semantic macro and each $@$ symbol are following siblings of the semantic macro node. That is the reason why we stored the number of parameters, variables and $@$ symbols in the lexicon files. Otherwise, the translator could not find the end of a semantic macro in the \gls*{pom-pt}.

@@

\Cref{fig:syntax-tree-usecase} visualizes the \gls*{pom-pt} of the Jacobi polynomial example from~\Cref{tab:JacobiP-usecase}. Because of the differences to expression trees, a backward conversion of the \gls*{pom-pt} to a string representation can be difficult, especially for finding necessary or unnecessary parentheses. Therefore we create the \gls*{teo}.

With these tools, we can translate a \LaTeX{} expression by translating the \gls*{pom-pt} node by node and perform group or reordering operations for some special cases. The algorithm is realized in a simple recursive structure. The Algorithm~\ref{alg:simple-translation} explains the translation process. Whenever the algorithm finds a leaf, it can translate this single term. If the node is not a leaf, it starts to translate all children of the node recursively.

\begin{algorithm}[t]
\caption{Abstract translation algorithm to translate MLP-Parse trees.}\label{alg:translation}
	\begin{algorithmic}[1]
	\Require Root $r$ of a POM-Parse tree $T$. List \textit{following\_siblings} with the following siblings of $r$. The list can be empty.
	\Procedure{abstract\_translator}{$r$, \textit{following\_siblings}}
	\If{$r$ is leaf}
		\State {\scriptsize TRANSLATE\_LEAF}($r$, \textit{following\_siblings});\label{line:transLeaf}
	\Else
		\State \textit{siblings} $ = r$.getChildren(); \Comment{\textit{siblings} is a list of children}
		\State {\scriptsize ABSTRACT\_TRANSLATOR}(\textit{siblings}.removeFirst(), \textit{siblings});\label{line:transSeq}
	\EndIf
	\If{\textit{following\_siblings} is not empty}\label{line:ifNotEmpty}
		\State $r =$ \textit{following\_siblings}.removeFirst();
		\State {\scriptsize ABSTRACT\_TRANSLATOR}($r$, \textit{following\_siblings});
	\EndIf
	\EndProcedure
	\end{algorithmic}
\end{algorithm}
%\vspace{-20px}

This approach is not completely feasible since the algorithm needs to look ahead and check the following siblings in some cases, e.g., in the case of a semantic macro with arguments (see~\Cref{fig:syntax-tree-usecase}). The {\footnotesize \verb|TRANSLATE_LEAF|} function needs to know the following siblings of the current node. Algorithm~\ref{alg:translation} explains a more detailed but abstract version of the final algorithm of the presented translation tool.

If the root $r$ is a leaf, it still can be translated as a leaf. Eventually, some of the following siblings are needed to translate $r$. The list of \textit{following\_siblings} in Line~\ref{line:transLeaf} might be reduced to avoid multiple translations for one node. If $r$ is not a leaf, it contains one or more children. Therefore, we can call the {\footnotesize \verb|ABSTRACT_TRANSLATOR|} recursively for the children. Once we have translated $r$, we can go a step further and translate the next node. Line~\ref{line:ifNotEmpty} checks if there are following siblings left and calls the {\footnotesize \verb|ABSTRACT_TRANSLATOR|} recursively in such cases. Translated expressions are stored by the \gls*{teo} object. Algorithm~\ref{alg:translation} is a simplified version of the translator process. The \Cref{line:transLeaf,line:transSeq} process the translations for each node. \Cref{tab:allTypesTable} gives an overview of all the different node types the root $r$ can be. A more detailed explanation of the types can be found in~\parencite{POM-Tagger}.

The \gls*{bnf} grammar defines some basic grammatical rules for generic \LaTeX{} macros, such as for \verb|\frac|, \verb|\sqrt|. Therefore, there is a hierarchical structure for those symbols similar to the structure in expression trees. As already mentioned, some of these types can be translated directly, such as Greek letters, while others are more complex, such as semantic \LaTeX{} macros. Therefore, the translators delegate the translation to specialized subtranslators. This delegation process is implemented in \Cref{line:transLeaf,line:transSeq} of Algorithm~{alg:translation}. The Subsection~\ref{sec:subtranslators} discusses these classes in more detail.

This approach seems to work for most of semantic \LaTeX{} expressions. However, there is one problem with this straight forward approach which was mentioned shortly mentioned in~\Cref{sec:problems}. There are many different types of notations used to represent formulae. \Cref{tab:notations} illustrates the simple expression $(a+b)x$ in different notations. The \gls*{pol}\footnote{Also known as \textit{Warsaw Notation} or \textit{prefix notation}} (hereafter called prefix notation) places the operator to the left of/before its operands. The \gls*{rpol}\footnote{Also known as \textit{postfix notation}} (hereafter called postfix notation) does the opposite and places the operator to the right of/after its operands. The infix notation is commonly used in arithmetic and places the operator between their operands, which only makes sense as long as the operator is a binary operator.

\begin{wrapfigure}{r}{0.38\textwidth}
\vspace{-10pt}
\begin{minipage}{0.38\textwidth}
\center
\begin{tabular}{lc}
	\hline
	Notation & Expression \\
	\hline
	Infix & $(a+b) \cdot x$\\
	Prefix & $\cdot + a\ b\ x$\\
	Postfix & $a\ b + x\ \cdot$\\
	Functional & $\cdot(+(a, b), x)$\\
	\hline
\end{tabular}
\vspace{-5pt}
\caption{The mathematical expression '$(a+b) \cdot x$' in infix, prefix, postfix and functional notation.}
\label{tab:notations}
\end{minipage}
\vspace{-5pt}
\end{wrapfigure}

@@

Most programming languages (and \gls*{cas} as well) internally use prefix or postfix notation and do not mix the notations in one expression, since it is more convenient to parse expressions in uniform notations. However, the common practice in science is to use mixed notations in expressions. Since the \gls*{pom} has rarely implemented mathematical grammatical rules yet, it takes the input as it is and does not build an expression tree. Therefore, it parses all four examples from~\Cref{tab:notations} to four different \gls*{pom-pt}s rather than to one unique expression tree. In general, this is not a problem for our translation process since most \gls*{cas} are familiar with the most common notations. Therefore, the translator does not need to know that $a$ and $b$ are the operands of the binary operator '$+$' in '$a+b$.' The translator could simply translate the symbols in '$a+b$' in the same order as they appear in the expression and the \gls*{cas} would understand it. However, this simple approach generates two problems.
\begin{enumerate}%[label=\arabic*)]
\item \label{prob:1} The translated expression is only syntactically correct if the input expression was syntactically correct.
\item \label{prob:2} We cannot translate expressions to a \gls*{cas} which uses a non-standard notation.
\end{enumerate}

@@

Nonetheless, changing a notation could also solve ambiguities in some situations. Consider the two ambiguous examples in~\Cref{tab:amb_ex}. While a scientist would probably just ask for the right interpretation of the first example, \Maple{} automatically computes the first interpretation. On the other hand, \LaTeX{} automatically disambiguate the first example by only recognizing the very next element (single symbols or sequence in curly brackets) for the superscript and therefore displays the second interpretation. The second example is already interpreted as the double factorial function of $n$, since this notation is the standard interpretation in science. We wrote the second interpretation as the standard way in science to make it even more obvious. However, surprisingly, \Maple{} computes the first interpretation again rather than the common standard interpretation.
\begin{table}[ht]
\centering
\begin{tabular}{lccc}
	\hline
	& Text Format Expression & First Interpretation & Second Interpretation\\
	\hline
	1: & \rule{0pt}{0.9\normalbaselineskip} $4\ \hat{\ }\ 2!$ & $4^{2!}$ & $4^2!$ \\
	2: & $n!!$ & $(n!)!$ & $(n)!!$\\
	\hline
\end{tabular}
\caption{Ambiguous examples of the factorial and double factorial function. One expression in a text format can be interpreted in different ways.}
\label{tab:amb_ex}
\end{table}

In most cases, parentheses can be used to disambiguate expressions. We used them in~\Cref{tab:amb_ex} to clarify the different interpretations in example 2. But sometimes, even parentheses cannot solve a mistaken computation. For example, there is no way to add parentheses to force \Maple{} to compute $n!!$ as the double factorial function. Even $(n)!!$ will be interpreted as $(n!)!$. Rather than using the exclamation mark in \Maple, one could also use the functional notation. For example, the interpretations $(2!)!$ and $(2)!!$ can be distinguished in \Maple{} by using \verb|factorial(factorial(2))| and \verb|doublefactorial(2)| respectively. We define the translations as follows:

@@

An exponent will be translated and shifted to the end of the translated semantic macro, since it is more common to write the exponent after the arguments of a function in \gls*{cas}. Therefore, the function translates and stores the exponent in Line~\ref{line:next_caret}. One could ask what happens when there is an exponent given before and after the arguments. The \verb|MacroTranslator| only translates the following siblings until each argument is translated. The first exponent will be shifted to the end. If right after the translated macro (with all arguments) follows another exponent, we interpret it as another exponent for the whole previous expression. In that case, it would be the macro with the first translated exponent. Table~\ref{tab:multi-expo} shows an example for the trigonometric cosine function with multiple exponents. 
\begin{table}[ht]
\centering
\begin{tabular}{lc}
\hline
Displayed As & \rule{0pt}{0.9\normalbaselineskip} $\cos^2(x)^2$ \\
Semantic \LaTeX & \verb|\cos^2@{x}^2| \\
Translated \Maple{} Expression & \verb|((cos(x))^(2))^2|\\
\hline
\end{tabular}
\caption{A trigonometric cosine function example with exponents before and after the argument.}
\label{tab:multi-expo}
\end{table}
